\section{Proposal}

In this paper we present a simplified proposal based on N3875~\cite{n3875}.

The core idea of this proposal is to allow a class to have a data member of
unspecified size and to provide mechanisms to separate size specification
from the rest of construction.

\subsection{Arrays of run-time bound as data members}

We propose a new syntax for introducing array data members of unspecified size.
The size will be specified at construction time.

\begin{lstlisting}
class myvec {
public:
  // ...
private:
  int sz;
  double[] v; //rtb array
};
\end{lstlisting}

\subsection{Constructors}

A \emph{run-time size bound class} needs to specify the size of its array data
members upon construction.

This paper proposes a new type of constructor: an array constructor. An array
constructor specifies the size of the array data member of unspecified size and
my forward a call to the corresponding non-array constructor. The array
constructor must always be inline, while the corresponding non-array
constructor does not need to be inline.

\begin{lstlisting}
class myvec {
public:
  myvec[]() : v[2], myvec{} {} // Array constructor must be inline
  myvec[](int n) : v[n], myvec{n} {} // Array constructor must be inline

  myvec(); // Need not be inline
  myvec(int n); // Need not be inline

  // ...
private
  int sz;
  double[] v;
};

myvec::myvec() : sz{2} { /* ... */ }
myvec::myvec(int n) : sz{n} { /* ... */ }
\end{lstlisting}

\subsection{Single constructor}

An array constructor can be used as single constructor instead of forwarding to a
non-array constructor.

\begin{lstlisting}
class myvec {
public:
  myvec[](int n) : sz{n}, v[n] {} // Needs to be inline

  // ...
private:
  int sz;
  double[] v;
};
\end{lstlisting}

Please, note that not using forwarding construction imposes the additional requirement
of being able to inline the constructor. When this is no desirable, still two constructors
(an array version and a non-array version) should be used.



\subsection{Implementing \emph{bs\_array}}

The proposed solution allows the implementation of \cppid{bs\_array}
(where bs stands for \emph{bike shed}) as a \emph{pure-library} solution.

\begin{lstlisting}
template <class T>
class bs_array {
public:
  using value_type = T;

  bs_array[](int n) : v[n], bs_array{n} {}
  bs_array(int n) : sz{n} {}

  bs_array(const bs_array & a) = default;
  bs_array(bs_array &&) = delete;

  // Unchecked access
  T & operator[](int i) { return v[i]; }
  const T & operator[](int i)  const { return v[i]; }

  // Range checked access
  T & at(int i);
  const T & at(int i) const;

  // begin/end
  T * begin() { return v; }
  const T * begin() const { return v; }
  T * end() { return v+sz; }
  const T * end() { return v+sz; }

  int size() const { return sz; }
  T * data() { return v; }

private:
  int sz;
  T[] v;
};
\end{lstlisting}

I propose that a class like this be standardized. While we are not proposing a
specific name for such class, alternate names could be \cppid{auto\_array},
\cppid{stack\_array}, \cppid{dynarray} or even \cppid{autodynarray}.

\subsection{Run-time size bound objects as data members}

A class that has one member of an array type of unspecified size is a
\emph{run-time size bound object} and its size cannot be determined at compile
time as the object size depends on the array size.

A class that has a run-time size bound data member becomes itself also into a
run-time size bound class. Such a class may have an array constructor.

\begin{lstlisting}
class A {
public:
  A[](int n) : sz{n}, v[n] {}
private:
  int sz;
  double[] v;
};

class B {
public:
  B[](int n) : x{n*1.5}, a{n} {}
private:
  double x;
  A a;
};
\end{lstlisting}

In general, a class can only have a run-time size bound data member (or 
an array of unspecified size data member) as its last data member. 

Similar restrictions apply to inheritance.

\begin{lstlisting}
class A {
public:
  A[](int n) : sz{n}, v[n] {}
private:
  int sz;
  double[] v;
};

class B : public A {
public:
  B[](int n) : x{n*1.5}, A{n} {}
private:
  double x;
};
\end{lstlisting}

A class inheriting from a run-time sized class cannot have subclasses that add new data members.

In case of multiple inheritance, only the most right base class can be a run-time sized class.

\begin{lstlisting}
class A {
public:
  A[](int n) : sz{n}, v[n] {}
private:
  int sz;
  double[] v;
};

class B { /* ... */ };

class D1 : public A, public B { /* ... */ }; // Ill-formed
class D2 : public B, public A { /* ... */ }; // Ill-formed
\end{lstlisting}

\subsubsection{Contexts of use}

A run-time size bound object can only be used as an automatic variable.
In particular, this paper proposes to ban any use implying direct or indirect
use of dynamic memory. Besides an array of such objects is also forbidden.

\begin{lstlisting}
class A {
public:
  A[](int n) : v[n], A{n} {}
  A(int n) : sz{n} {}
private:
  int sz;
  double[] v;
};

void f() {
  A a{4}; // OK;
  A[4] v; // Error
  A * p = new A{4}; // Error
  vector<A> w; // Error
}
\end{lstlisting}

\subsubsection{\emph{sizeof}}

Operator \cppkey{sizeof} is not supported on any type or object of a \emph{run-time size bound class}.

Firstly, on types \cppkey{sizeof} makes not much sense as different object of such types could require
different amounts of storage, due to a member of unspecified size. Secondly, implementing \cppkey{sizeof}
on objects would lead to different objects of the same type having a different size, and requiring
additional implementation overhead.

Although, there might be some implementation workarounds there is at least on national body
that has expressed very strong opposition to this feature.

\subsection{Parameter passing and return values}

Parameters of \emph{run-time size bound classes} can only be passed as parameters to functions (or returned)
when such passing mode does not imply an object copy. This restriction avoids problems with size determination.

\begin{lstlisting}
void f(bs_array<int> & a); // OK
bs_array<double> & g(const bs_array<int> & a); // OK
bs_array<char> h(); // Ill-formed
void i(bs_array<long> a); // Ill-formed
\end{lstlisting}

\subsection{Inlining versus non inlining}

In contrast with N3875, this proposal does not require constructors to be
generally inlined. The only constructor that needs to be inlined is the
array constructor. Others may be inlined or not.

\subsection{Size determination}
\label{sec:inline-size-det}

The current proposal implies that the size of any \emph{run-time bound array
data member} can be derived from the environment where the inline constructor is
defined.

In particular, this proposal does not allow the following example (slightly
modified from an example provided by Lawrence Crowl).

\begin{lstlisting}
// a.h
struct A {
  bs_array<double> storage;
  A[](int n);
}; // Error A needs an inlined array constructor

// a.cc
  extern int config;
  A::A[](int n) : storage{n*config} {} // Not inline: Ill-formed

// b.cc
void f() {
  A x{10}; // Ill-formed.
  //...
}
\end{lstlisting}

As \cppkey{struct} \cppid{A} has a data member which is a \emph{run-time size bound class}
(the \cppid{bs\_array}), it is considered itself to be also a \emph{run-time size
bound class} and it needs an inlined array constructor.

